
\section{Problems and Datasets}

3 Datasets will be used to evaluate the ByteNet and semi-supervised ByteNet models. The WMT Translation Task provides a large corpus of text in form of diffrent datasets. Two datasets are used in this thesis, the Europarl v7 dataset and the WMT NewTest dataset. For validating the implementation and discussing the model a very simple synthetic dataset is also used, this dataset is named ``Synthetic Digits''.

\subsection{WMT Translation Task}

Each year WMT holds a conference on machine translation, this works as a series of workshops and competitions on diffrent translation tasks. One of these translation tasks are on news translation \cite{wmt16}. The task is to translate paragraphs from news articles to another language. The translations are evaluated on the WMT NewsTest dataset, that are from news articles collected over a specific year. The news translation task is the primary translation task that NMT papers evaluates their models on \cite{bytenet, bahdanau-2015-nmt, sutskever-2014-nmt}.

\begin{table}[H]
\centering
\begin{tabular}{r|p{5cm} p{5cm}}
	obs. & source & target\\ \hline
     0 & Die Premierminister Indiens und Japans trafen sich in Tokio. & India and Japan prime ministers meet in Tokyo \\
     1 & Pläne für eine stärkere kerntechnische Zusammenarbeit stehen ganz oben auf der Tagesordnung. & High on the agenda are plans for greater nuclear co-operation. \\
     2 & Berichten zufolge hofft Indien darüber hinaus auf einen Vertrag zur Verteidigungszusammenarbeit zwischen den beiden Nationen. & India is also reportedly hoping for a deal on defence collaboration between the two nations.
\end{tabular}
\caption{Example form the WMT NewsTest 2015 de-en dataset.}
\end{table}

The WMT NewsTest is about 3000 sentences each year, this is not enough data to build a good translation model. To that end WMT provides additional datasets for training, most importantly is the Europarl v7 dataset. The Europarl dataset contains high quality translations of various of documents from the European Parlement \cite{europarl}. While this dataset is a huge high quality dataset it is also heavily biased, as some word grams like ``The European Parlement'' appears much more often that in the NewsTest dataset.

\begin{table}[H]
\centering
\begin{tabular}{r|p{5cm} p{5cm}}
	obs. & source & target\\ \hline
        0 & Wir sind nach wie vor der Ansicht, daß der wirtschaftliche und soziale Zusammenhalt ein zentrales Ziel der Union ist. & We still feel that economic and social cohesion is one of the Union' s fundamental objectives. \\
        1 & Den Kommissar möchte ich auffordern, in diesen beiden Bereichen tätig zu werden und uns dabei mit einzubinden. & These are two areas of action which I invite the Commissioner to set up and in which I would ask him to involve us. \\
        2  & Dies zwingt das Europäische Parlament, den Herrn Kommissar und die Kommission zu entschlossenem strategischen Handeln. & This fact means that the European Parliament, the Commissioner and the Commission must act decisively and strategically.
\end{tabular}
\caption{Example form the Europarl v7 de-en dataset.}
\end{table}

\subsection{Synthetic Digits}

While WMT Translation Task is good problem to measure a translation model on, it is a complex problem that takes a long time for a neural network to learn. It is also not obvious if a model will be able to solve this problem at all. In particular the Semi-Supervised ByteNet model has never been attempted before, thus it is meaning full to validate that it can solve a simple problem before attempting more a more complex problem like the WMT Translation Task.

The synthetic digits dataset is simply a uniformly random sequence of integers. In the source each digit is spelled out in english, in the target each digit is represented as a symbol. The sequence length is uniformly randomly chosen to be either 2 or 3 digits long.
\begin{table}[H]
\centering
\begin{tabular}{r|p{5cm} p{5cm}}
	obs. & source & target\\ \hline
	0 & zero two & 02 \\
    1 & four eight & 48 \\
    2 & eight four four & 844
\end{tabular}
\caption{Example form the Synthetic Digits dataset generator.}
\end{table}
